---
title: "Statistics 7.2.1 Exercises 1 and 2"
subtitle: "Standard Error and Normal Distributions"
author: "Lucas Liona"
format:
  pdf:
    toc: true
    fig-width: 10
    fig-height: 8
  html:
    toc: true
    code-fold: show
    fig-width: 10
    fig-height: 8
execute:
  echo: true
---

## Exercise 1: Sample Size and Standard Error

We know that $\text{sd}(\bar{X}) = \frac{\sigma}{\sqrt{n}}$, where $\sigma$ is the population standard deviation and $n$ is the sample size. Also, $\text{sd}(\bar{X})$ decreases proportionally to $\frac{1}{\sqrt{n}}$.

Given information:
- Initial sample size is 10 observations
- At this sample size, $\text{sd}(\bar{X}) = 9$

Let's solve for how many additional observations are needed in each scenario.

### Part (a): Reducing $\text{sd}(\bar{X})$ to 4.5

We need to find the new sample size $n_2$ such that $\text{sd}_2(\bar{X}) = 4.5$.

Using the relationship between standard errors and sample sizes:
$\frac{\text{sd}_2(\bar{X})}{\text{sd}_1(\bar{X})} = \sqrt{\frac{n_1}{n_2}}$

$\frac{4.5}{9} = \sqrt{\frac{10}{n_2}}$

$\left(\frac{4.5}{9}\right)^2 = \frac{10}{n_2}$

$\frac{20.25}{81} = \frac{10}{n_2}$

$n_2 = \frac{10 \times 81}{20.25} = \frac{810}{20.25} = 40$

Since we already have 10 observations, we need an additional 30 observations.

```{r}
# Verify calculation
n1 <- 10
sd1 <- 9
sd2 <- 4.5

n2 <- n1 * (sd1/sd2)^2
additional_obs <- n2 - n1

cat("New sample size needed:", n2, "\n")
cat("Additional observations needed:", additional_obs, "\n")
```

### Part (b): Reducing $\text{sd}(\bar{X})$ to 3

Using the same approach:
$\frac{3}{9} = \sqrt{\frac{10}{n_2}}$

$\left(\frac{3}{9}\right)^2 = \frac{10}{n_2}$

$\frac{9}{81} = \frac{10}{n_2}$

$n_2 = \frac{10 \times 81}{9} = \frac{810}{9} = 90$

Since we already have 10 observations, we need an additional 80 observations.

```{r}
sd2 <- 3
n2 <- n1 * (sd1/sd2)^2
additional_obs <- n2 - n1

cat("New sample size needed:", n2, "\n")
cat("Additional observations needed:", additional_obs, "\n")
```

### Part (c): Reducing $\text{sd}(\bar{X})$ to 1

Using the same approach:
$\frac{1}{9} = \sqrt{\frac{10}{n_2}}$

$\left(\frac{1}{9}\right)^2 = \frac{10}{n_2}$

$\frac{1}{81} = \frac{10}{n_2}$

$n_2 = \frac{10 \times 81}{1} = 810$

Since we already have 10 observations, we need an additional 800 observations.

```{r}
sd2 <- 1
n2 <- n1 * (sd1/sd2)^2
additional_obs <- n2 - n1

cat("New sample size needed:", n2, "\n")
cat("Additional observations needed:", additional_obs, "\n")
```

## Exercise 2: Normal Distribution and Probability Calculation

### Problem Setup

Given information:
- Monthly profits ($X$) follow Normal distribution: $X \sim \text{Normal}(\mu = 10, \sigma = 3.5)$ million dollars
- Average is calculated over a six-month period: $\bar{X}$ 
- Company pays dividend if $\bar{X} > 8.5$ million
- Need to find probability that company will pay a dividend

### Finding the Distribution of $\bar{X}$

When individual observations follow a Normal distribution, the sampling distribution of the mean also follows a Normal distribution:

$\bar{X} \sim \text{Normal}(\mu_{\bar{X}}, \sigma_{\bar{X}})$

Where:
- $\mu_{\bar{X}} = \mu = 10$ million (the mean of $\bar{X}$ equals the population mean)
- $\sigma_{\bar{X}} = \frac{\sigma}{\sqrt{n}} = \frac{3.5}{\sqrt{6}} = \frac{3.5}{2.45} \approx 1.429$ million

```{r}
mu <- 10       # Population mean
sigma <- 3.5   # Population standard deviation
n <- 6         # Sample size (6 months)

mu_x_bar <- mu
sigma_x_bar <- sigma / sqrt(n)

cat("Mean of sampling distribution:", mu_x_bar, "million\n")
cat("Standard deviation of sampling distribution:", sigma_x_bar, "million\n")
```

### Calculating the Probability of Paying Dividend

We need to find $P(\bar{X} > 8.5)$

First, we standardize to find the z-score:
$z = \frac{8.5 - 10}{1.429} \approx -1.05$

Then, we find the probability:
$P(\bar{X} > 8.5) = P(Z > -1.05) = 1 - P(Z \leq -1.05)$

```{r}
threshold <- 8.5
z_score <- (threshold - mu_x_bar) / sigma_x_bar
probability <- 1 - pnorm(z_score)

cat("Z-score:", z_score, "\n")
cat("Probability of paying dividend:", probability, "\n")
cat("Probability as percentage:", probability * 100, "%\n")
```

The probability that the company will pay an annual dividend is approximately 0.853 or 85.3%.