---
title: "Probability with R - Exercise 2.1"
subtitle: "Skewness"
author: "Lucas Liona"
format:
  pdf:
    toc: true
    fig-width: 8
    fig-height: 6
  html:
    toc: true
    code-fold: show
    fig-width: 8
    fig-height: 6
execute:
  echo: true
---

## Problem

1.  Write the skewness program and use it to calculate the skewness coefficient of the four examination subjects in results.txt. What can you say about these data?

2.  Pearson has given an approximate formula for the skewness, which is easier to calculate than the exact formula:

    $\text{skew} \approx \frac{3(\text{mean} - \text{median})}{\text{sd}}$

Write a program to calculate this and apply it to the data in results.txt. Is it a reasonable approximation?

3.  For the class of 50 students of computing detailed in Exercises 1.1, use R to

<!-- -->

(a) obtain the summary statistics for each gender and for the entire class;
(b) calculate the deciles for each gender and for the entire class;
(c) obtain the skewness coefficient for the females and for the males.

<!-- -->

4.  It is required to estimate the number of message buffers in use in the main memory of the computer system at Roantree Retailers Ltd. To do this, 20 programs were run and the numbers of message buffers in use were found to be

    ```         
    141, 146, 157, 151, 152, 140, 142, 156, 150, 140 
    139, 135, 143, 146, 146, 152, 140, 136, 149, 148
    ```

Calculate the average number of buffers used. What is the standard deviation?

Would you say these data are skewed?

## Skewness Coefficient

text/explain

```{r}
skew <- function(x) {
sum2 <- sum((x-mean(x))^2)
sum3 <- sum((x-mean(x))^3)
skew <- (sqrt(length(x))* sum3)/(sum2^(1.5))
skew}

skew()

```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
