---
title: "Statistics 8.4 Exercise 2"
subtitle: "95% Confidence Interval"
author: "Lucas Liona"
format:
  pdf:
    toc: true
    fig-width: 8
    fig-height: 6
  html:
    toc: true
    code-fold: show
    fig-width: 8
    fig-height: 6
execute:
  echo: true
---

## Problem
Calculate a 95% confidence interval for the difference for the true mean testosterone levels for 1- to 30-per-day smokers and 31- to 70-per-day smokers from the data given in the Exercises for Section 7.5, problem 4.

## Solution

First, let's establish the summary statistics from the provided data:

- 1-30 cigarettes/day: $\bar{x}_1 = 715.6$ ng/dL, $s_1 = 248.0$ ng/dL, $n_1 = 31$  
- 31-70 cigarettes/day: $\bar{x}_2 = 795.1$ ng/dL, $s_2 = 305.3$ ng/dL, $n_2 = 28$

For the 95% confidence interval, we need the appropriate t-value with degrees of freedom equal to the minimum of $(n_1 - 1)$ and $(n_2 - 1)$.

```{r}
# Calculate degrees of freedom
df <- min(31 - 1, 28 - 1)
df

# Find t-value for 95% CI
t_value <- qt(0.975, df)
t_value
```

The standard error of the difference between means is:

```{r}
# Calculate standard error
se_diff <- sqrt((248.0^2/31) + (305.3^2/28))
se_diff
```

Now we can calculate the 95% confidence interval:

```{r}
# Calculate difference in means
mean_diff <- 715.6 - 795.1
mean_diff

# Calculate margin of error
margin_error <- t_value * se_diff

# Calculate confidence interval
lower_bound <- mean_diff - margin_error
upper_bound <- mean_diff + margin_error

c(lower_bound, upper_bound)
```

## Interpretation

With 95% confidence, the true difference in mean testosterone levels between 1-30 per day smokers and 31-70 per day smokers falls between `r round(lower_bound, 1)` ng/dL and `r round(upper_bound, 1)` ng/dL.

Since this interval includes zero, we cannot conclude there is a statistically significant difference in testosterone levels between these two groups of smokers at the 95% confidence level.