---
title: "Statistics 6 Review Exercise 2"
subtitle: "Analysis of Cancer Patient Survival Times"
author: "Lucas Liona"
format:
  pdf:
    toc: true
    fig-width: 10
    fig-height: 8
  html:
    toc: true
    code-fold: show
    fig-width: 10
    fig-height: 8
execute:
  echo: true
---

## Problem

A medical trial was conducted to investigate whether a new drug extended the life of a patient who had lung cancer. The survival times (in months) for 38 cancer patients who were treated with the drug are as follows:

5, 9, 10, 13, 14, 18, 18, 19, 20, 22, 25, 25, 25, 27, 28, 30, 30, 33, 36, 38, 39, 39, 40, 41, 41, 43, 44, 44, 45, 46, 46, 49, 50, 50, 54, 54, 59

The sample mean is approximately 31.1 months and the standard deviation is approximately 16.0 months.

## Part (a): Normality Assessment

Before performing any calculations assuming normality, we should first assess whether the data follows a normal distribution.

```{r}
# Input the data
survival_times <- c(5, 9, 10, 13, 14, 18, 18, 19, 20, 22, 25, 25, 25, 27, 28, 30, 30, 33, 36, 38, 39, 39, 40, 41, 41, 43, 44, 44, 45, 46, 46, 49, 50, 50, 54, 54, 59)

# Parameters for normal distribution
mu <- 31.1
sigma <- 16.0

# Perform Shapiro-Wilk test for normality
shapiro_test <- shapiro.test(survival_times)
shapiro_test

# Create a comprehensive normality assessment plot
par(mfrow = c(2, 2))

# 1. Histogram with normal curve
hist(survival_times, freq = FALSE, 
     main = "Histogram with Normal Curve",
     xlab = "Survival Time (months)", 
     col = "lightblue",
     breaks = "Sturges")
curve(dnorm(x, mean = mu, sd = sigma), 
      add = TRUE, col = "red", lwd = 2)

# 2. Q-Q Plot
qqnorm(survival_times, main = "Normal Q-Q Plot")
qqline(survival_times, col = "red")

# 3. Boxplot
boxplot(survival_times, main = "Boxplot",
        ylab = "Survival Time (months)",
        col = "lightblue")

# 4. Density plot
plot(density(survival_times), 
     main = "Density Plot",
     xlab = "Survival Time (months)")
curve(dnorm(x, mean = mu, sd = sigma), 
      add = TRUE, col = "red", lwd = 2)
```

## Part (b): Normal Distribution Calculations

Despite the non-normality suggested by the Shapiro-Wilk test (p-value = `r round(shapiro_test$p.value, 4)`), we'll proceed with the normal distribution calculations as requested, but with appropriate caveats.

### (i) Probability of Surviving No More Than One Year

```{r}
# Calculate P(X ≤ 12)
prob_less_than_12 <- pnorm(12, mean = mu, sd = sigma)
prob_less_than_12
```

The probability that a patient survives for no more than one year (12 months) is approximately `r round(prob_less_than_12 * 100, 2)`%.

### (ii) Proportion Surviving Between One and Two Years

```{r}
# Calculate P(12 < X ≤ 24)
prob_less_than_24 <- pnorm(24, mean = mu, sd = sigma)
prob_between_12_and_24 <- prob_less_than_24 - prob_less_than_12
prob_between_12_and_24
```

The proportion of patients who survive between one year (12 months) and two years (24 months) is approximately `r round(prob_between_12_and_24 * 100, 2)`%.

### (iii) 80th Percentile of Survival Times

```{r}
# Calculate the 80th percentile
percentile_80 <- qnorm(0.8, mean = mu, sd = sigma)
percentile_80
```

The highest number of months that 80% of patients survive is approximately `r round(percentile_80, 1)` months.

### (iv) Central 80% of Survival Times

```{r}
# Calculate the 10th and 90th percentiles
percentile_10 <- qnorm(0.1, mean = mu, sd = sigma)
percentile_90 <- qnorm(0.9, mean = mu, sd = sigma)

# Output the interval
central_80_percent <- c(percentile_10, percentile_90)
central_80_percent
```

The central 80% of survival times falls between `r round(percentile_10, 1)` months and `r round(percentile_90, 1)` months.

## Part (c): Non-parametric Analysis

Given the non-normality of the data, let's also perform a non-parametric analysis for comparison:

```{r}
# Calculate empirical probabilities
empirical_less_than_12 <- mean(survival_times <= 12)
empirical_between_12_24 <- mean(survival_times > 12 & survival_times <= 24)

# Calculate empirical percentiles
empirical_80th <- quantile(survival_times, 0.8)
empirical_10th <- quantile(survival_times, 0.1)
empirical_90th <- quantile(survival_times, 0.9)

# Create comparison table
comparison <- data.frame(
  Measure = c("P(X ≤ 12)", "P(12 < X ≤ 24)", "80th Percentile", "10th Percentile", "90th Percentile"),
  Normal = c(prob_less_than_12, prob_between_12_and_24, percentile_80, percentile_10, percentile_90),
  Empirical = c(empirical_less_than_12, empirical_between_12_24, empirical_80th, empirical_10th, empirical_90th)
)

knitr::kable(comparison, digits = 3, 
             caption = "Comparison of Normal and Empirical Estimates")
```

## Conclusions

1. **Normality Assessment**: The Shapiro-Wilk test and visual inspections suggest that the survival times do not follow a normal distribution. The data appears to be right-skewed, which is common for survival data.

2. **Comparison of Methods**: The normal distribution calculations provide reasonable approximations for some measures but show notable differences for others, particularly in the tails of the distribution. The empirical (non-parametric) estimates are more accurate for this dataset.

3. **Practical Implications**: When working with survival data, it's often more appropriate to use non-parametric methods or specialized survival analysis techniques rather than assuming normality. The normal distribution calculations should be interpreted with caution, especially for extreme percentiles.

4. **Recommendations**: For more accurate analysis of this type of data, consider using:
   - Kaplan-Meier survival curves
   - Log-normal or Weibull distributions
   - Non-parametric confidence intervals
   - Cox proportional hazards models for more complex analyses