---
title: "Statistics 3.2.2 Exercise 2"
subtitle: "Waist-to-Hip Ratios Men/Women"
author: "Lucas Liona"
format:
  pdf:
    toc: true
    fig-width: 8
    fig-height: 6
  html:
    toc: true
    code-fold: show
    fig-width: 8
    fig-height: 6
execute:
  echo: true
---

## Problem Description

Waist-to-hip ratios of men and women sampled from the workforce-study database are provided. The data includes different age groups (≤45 and >45) for both males and females.

## Data Loading and Preparation

```{r}
# Load required libraries
library(ggplot2)
library(dplyr)

# Read the data
waist_data <- read.table("waists.txt", header = TRUE, sep = "\t")

# Check the structure of our data
str(waist_data)

# View summary statistics
summary(waist_data)
```

## (a) Box Plots

Let's create box plots to compare the four groups by gender and age group:

```{r}
# Create box plots for each group
ggplot(waist_data, aes(x = agegp, y = ratios, fill = sex)) +
  geom_boxplot() +
  facet_wrap(~ sex) +
  labs(title = "Waist-to-Hip Ratios by Gender and Age Group",
       x = "Age Group",
       y = "Waist-to-Hip Ratio") +
  theme_minimal() +
  scale_fill_manual(values = c("male" = "lightblue", "feml" = "pink"))
```

```{r}
# Alternative: Side-by-side box plots
par(mfrow = c(1, 2))

# Plot for males
male_data <- waist_data[waist_data$sex == "male", ]
boxplot(ratios ~ agegp, data = male_data, 
        main = "Males", 
        xlab = "Age Group", 
        ylab = "Waist-to-Hip Ratio",
        ylim = c(0.7, 1.1))

# Plot for females
female_data <- waist_data[waist_data$sex == "feml", ]
boxplot(ratios ~ agegp, data = female_data, 
        main = "Females", 
        xlab = "Age Group", 
        ylab = "Waist-to-Hip Ratio",
        ylim = c(0.7, 1.1))
```

## (b) Analysis of Results

Examining the box plots reveals several important patterns:

### Gender Differences:
- Women consistently have lower waist-to-hip ratios than men across both age groups.
- Male ratios typically range from about 0.85 to 1.0, while female ratios range from about 0.7 to 0.9.

### Age Effects:
- For both genders, the older age group (>45) shows slightly higher waist-to-hip ratios on average.
- The age effect appears more pronounced in women, with a clearer separation between the age groups.

### Variability:
- All groups show relatively similar spreads (interquartile ranges).
- The data appears approximately symmetric for all groups with a few outliers in each.

The clear separation between male and female distributions suggests that gender is a strong predictor of waist-to-hip ratio. Age also appears to be a factor, though less pronounced than gender.

## Statistical Summary

Let's calculate summary statistics for each group:

```{r}
# Group summary statistics
waist_data %>%
  group_by(sex, agegp) %>%
  summarise(
    n = n(),
    mean = mean(ratios),
    median = median(ratios),
    sd = sd(ratios),
    min = min(ratios),
    max = max(ratios)
  ) %>%
  knitr::kable(digits = 3, caption = "Summary Statistics by Group")
```

## Conclusion

The analysis confirms that both gender and age influence waist-to-hip ratios:

1. **Gender is the primary factor**: Men consistently have higher ratios than women, likely reflecting biological differences in fat distribution patterns.

2. **Age has a secondary effect**: Both genders show increased ratios with age, possibly due to changes in body composition over time.

These differences are important for health assessments, as waist-to-hip ratio is often used as an indicator of health risks associated with body fat distribution.